---
title: "Johns Hopkins COVID 19 Data Analysis"
author: "C. Nguyen"
date: "`r Sys.Date()`"
output: html_document
---

```{r load-packages, include=FALSE}
library(tidyverse)
library(lubridate)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## COVID19 Data Analysis
Analysis of COVID19 data for the US and globally. Looking in to case and death rates in different areas.

### Read in Data
Start by reading in the 4 main csv files
```{r get_jhu_data}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv","time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv")

urls <- str_c(url_in,file_names)

global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])

```

##Tidy up global data

Change values to cases per date and remove extra columns
```{r tidy_global_data}
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)
summary(global)
```

## Tidy up US data
Do same tidying of US data as was done for global data
```{r tidy_us_data}

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases")%>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "deaths")%>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)
  
US <- US %>%
   unite("Combined_Key",
         c(Province_State, Country_Region),
         sep = ", ",
         na.rm = TRUE,
         remove = FALSE)
global <- global %>%
   unite("Combined_Key",
         c(Province_State, Country_Region),
         sep = ", ",
         na.rm = TRUE,
         remove = FALSE)
```

Fill in missing population values in both dataframes
``` {r get_populations}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Combined_Key", "Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         Population, Combined_Key)

US <- US %>%
  left_join(uid, by = c("Combined_Key", "Province_State",
                        "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         Population)
```

## Visualization
Group by state
```{r group_data}
#group by state
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population), na.rm = TRUE) %>%
  mutate(deaths_per_mill = deaths *1000000 /
           Population) %>%
  select(Province_State, Country_Region, date, cases,
         deaths,
         deaths_per_mill, Population) %>%
  ungroup()
US_by_state <- na.omit(US_by_state)
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/
           Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population)%>%
  ungroup()
```

Plot COVID19 totals for US and Maryland
```{r plot_totals}
#plot
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10() +
  theme(legend.position="bottom") +
  labs(title = "COVID19 in the US", y = NULL)
  
state <- "Maryland"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10() +
  theme(legend.position="bottom") +
  labs(title = "COVID19 in Maryland", y = NULL)
  
max(US_totals$date)
max(US_totals$cases)
max(US_totals$deaths)
```

Look at number of new cases.

```{r analyze_data}
#number of new cases
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths"))+
  geom_point(aes(y = new_deaths, color =
                   "new_deaths"))+
  scale_y_log10() +
  theme(legend.position="bottom") +
  labs(title = "COVID19 in the US", y = NULL)

#max deaths per state

US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000*cases/population,
            deaths_per_thou = 1000*deaths/population) %>%
  filter(cases > 0, population > 0)

#get smallest 10
US_state_totals %>%
  slice_min(deaths_per_thou, n = 10)

#can also do slice_max
```

### Fit Data to Model
Compare deaths per thousands to cases per thousands.
Fit to model and plot.

```{r model_data}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

summary(mod)

#x_grid <- seq(1, 151)
#new_df <- tibble(cases_per_thou = x_grid)

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod)) %>% select(pred, deaths_per_thou, cases_per_thou, everything())

US_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")+
  labs(title = "Predicted Deaths per Thousand based on Cases per Thousand")
```

### Conclusion

I was able to fit the deaths per thousand and cases per thousand to a linear model which shows that the number of deaths increases with the number of cases. 

